---
title: "`r paste0('Testing affect regulation via computational modeling of real-time alcohol choices in daily life: ', params$outcome_label)`"
author: ""
date: "`r Sys.Date()`"
params:
  outcome: "choice_prop"
  outcome_label: "Choice Proportion"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: cosmo
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)
library(targets)
library(knitr)
library(kableExtra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ranger)
library(glmmLasso)

theme_set(theme_minimal(base_size = 13))

# Create figures directory
fig_dir <- file.path("figures", params$outcome)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

# Helper to save a ggplot as publication-quality JPG
save_plot <- function(p, filename, width = 9, height = 6, dpi = 300) {
  filepath <- file.path(fig_dir, paste0(filename, ".pdf"))
  ggsave(filepath, plot = p, width = width, height = height, units = "in",
         device = cairo_pdf, bg = "white")
}

# Helper to reliably print kable HTML (this gave me issues before)
print_kable <- function(x) {
  cat(as.character(x))
}
```

```{r config}
analyses <- list(
  list(name = "Stress Binary",  predictor = "stress_binary",
       store = "_targets_store_stress_binary"),
  list(name = "Stress State",   predictor = "stress_state",
       store = "_targets_store_stress_state"),
  list(name = "Negative Affect State", predictor = "NA_state",
       store = "_targets_store_na_state"),
  list(name = "Positive Affect State", predictor = "PA_state",
       store = "_targets_store_pa_state")
)

out <- params$outcome

lambda_seqs <- list(
  choice_prop = seq(100, 0, by = -5),
  median_rt   = seq(100, 0, by = -5),
  drift       = seq(100, 0, by = -5),
  alcbias     = seq(200, 0, by = -10),
  B           = seq(200, 0, by = -10)
)
```

```{r helper-functions}
get_component_info <- function(df) {
  pc_cols <- grep("^PC_", names(df), value = TRUE)
  umap_cols <- grep("^UMAP_", names(df), value = TRUE)
  list(n_pca = length(pc_cols), n_umap = length(umap_cols))
}

get_component_composition <- function(df, outcome, predictor, prefix = "PC_", n_top = 5) {
  comp_cols <- grep(paste0("^", prefix), names(df), value = TRUE)
  raw_cols <- setdiff(names(df),
                      c("pid", outcome, predictor,
                        grep("^PC_|^UMAP_", names(df), value = TRUE)))
  results <- list()
  for (comp in comp_cols) {
    cors <- sapply(raw_cols, function(v) cor(df[[comp]], df[[v]], use = "complete.obs"))
    top_idx <- order(abs(cors), decreasing = TRUE)[1:min(n_top, length(cors))]
    results[[comp]] <- data.frame(
      Variable = raw_cols[top_idx],
      Correlation = round(cors[top_idx], 3),
      stringsAsFactors = FALSE, row.names = NULL)
  }
  results
}

format_glmm_coefs <- function(glmm_result, variant_label) {
  coefs <- glmm_result$coefficients
  if (is.null(coefs)) return(NULL)
  data.frame(Term = names(coefs), Estimate = round(as.numeric(coefs), 4),
             stringsAsFactors = FALSE)
}

get_variant_label <- function(formula_obj) {
  f_text <- as.character(formula_obj)
  if (any(grepl("PC_", f_text))) return("PCA")
  if (any(grepl("UMAP_", f_text))) return("UMAP")
  return("Raw")
}

reorder_by_variant <- function(formulas, results_list) {
  labels <- sapply(formulas, get_variant_label)
  desired_order <- c("Raw", "PCA", "UMAP")
  ord <- match(labels, desired_order)
  list(formulas = formulas[order(ord)], results = results_list[order(ord)])
}

classify_glmm_terms <- function(term_names, predictor) {
  is_intercept <- term_names == "(Intercept)"
  is_focal <- term_names == predictor
  is_interaction_with_focal <- (
    grepl(paste0(":", predictor, "$"), term_names) |
    grepl(paste0("^", predictor, ":"), term_names)
  ) & !is_focal

  category <- rep("Covariate main effect", length(term_names))
  category[is_intercept] <- "Intercept"
  category[is_focal] <- "Focal predictor"
  category[is_interaction_with_focal] <- "Interaction with focal predictor"
  category
}

refit_ranger_with_importance <- function(forest_result, df_analysis, formula_obj) {
  rec <- forest_result$recommended.pars
  outcome_var <- as.character(formula_obj)[2]
  pred_vars <- setdiff(all.vars(formula_obj), outcome_var)
  model_data <- df_analysis[, c(outcome_var, pred_vars)]

  simple_formula <- as.formula(paste(outcome_var, "~ ."))
  ranger(simple_formula, data = model_data, num.trees = 500,
         mtry = rec$mtry, min.node.size = rec$min.node.size,
         sample.fraction = rec$sample.fraction, importance = "permutation")
}

compute_pdp <- function(model, data, var_name, grid_size = 25) {
  if (is.numeric(data[[var_name]])) {
    grid_vals <- seq(min(data[[var_name]], na.rm = TRUE),
                     max(data[[var_name]], na.rm = TRUE),
                     length.out = grid_size)
  } else {
    grid_vals <- unique(data[[var_name]])
  }
  pd <- sapply(grid_vals, function(val) {
    data_mod <- data
    data_mod[[var_name]] <- val
    mean(predict(model, data = data_mod)$predictions)
  })
  data.frame(value = grid_vals, yhat = pd, variable = var_name,
             stringsAsFactors = FALSE)
}

get_glmm_fitted <- function(model) {
  if (!is.null(model$y_hat)) return(as.numeric(model$y_hat))
  if (!is.null(model$fitted.values)) return(as.numeric(model$fitted.values))
  return(NULL)
}
```

# About This Report

This report presents results for **`r params$outcome_label`** (`r params$outcome`) across all four focal predictors: stress (binary), stress (state), negative affect (state), and positive affect (state).

For each predictor, the report is organized as follows:

- **Dimensionality reduction:** PCA and UMAP component composition tables, showing which raw variables load most strongly onto each component.
- **GLMM results (raw features):** Summary of results from GLMM model with raw variables.
- **Random forest results (raw features):** Summary of results from random forest model with raw variables.
- **Model comparison:** RMSE comparison table across all methods (GLMM, random forest, decision tree) and feature sets (raw, PCA, UMAP).

```{r results='asis'}
for (analysis in analyses) {

  cat("\n\n# ", analysis$name, " (", analysis$predictor, ")\n\n", sep = "")

  # ---- Load targets ----
  df_analysis <- tar_read_raw(paste0("df_analysis_", out), store = analysis$store)
  all_formulas <- tar_read_raw(paste0("all_formulas_", out), store = analysis$store)
  all_glmms <- tar_read_raw(paste0("all_glmms_", out), store = analysis$store)
  all_forests <- tar_read_raw(paste0("all_forests_", out), store = analysis$store)
  tab_fits <- tar_read_raw(paste0("tab_fits_", out), store = analysis$store)

  # Identify raw variant
  variant_labels <- sapply(all_formulas, get_variant_label)
  raw_idx <- which(variant_labels == "Raw")
  raw_formula <- all_formulas[[raw_idx]]
  raw_glmm <- all_glmms[[raw_idx]]
  raw_forest <- all_forests[[raw_idx]]

  predictor <- analysis$predictor
  lambdas <- lambda_seqs[[out]]
  outcome_var <- out
  y_var <- df_analysis[[out]]
  y_range <- range(y_var, na.rm = TRUE)

  # Prefix for figure filenames
  fig_prefix <- paste0(predictor, "_", out)

  # ===========================================================================
  # DIMENSIONALITY REDUCTION
  # ===========================================================================
  comp_info <- get_component_info(df_analysis)

  cat("## Dimensionality Reduction\n\n")
  cat("- **PCA components (80% variance, max 5):**", comp_info$n_pca, "\n")
  cat("- **UMAP components (matching PCA):**", comp_info$n_umap, "\n\n")

  cat("### PCA Component Composition (top 5 correlated variables)\n\n")
  pca_comp <- get_component_composition(df_analysis, out, predictor, prefix = "PC_")
  for (pc_name in names(pca_comp)) {
    cat("**", pc_name, "**\n\n", sep = "")
    print_kable(
      kable(pca_comp[[pc_name]], row.names = FALSE, format = "html") %>%
        kable_styling(bootstrap_options = c("striped", "condensed"),
                      full_width = FALSE, font_size = 12)
    )
    cat("\n\n")
  }

  cat("### UMAP Component Composition (top 5 correlated variables)\n\n")
  umap_comp <- get_component_composition(df_analysis, out, predictor, prefix = "UMAP_")
  for (umap_name in names(umap_comp)) {
    cat("**", umap_name, "**\n\n", sep = "")
    print_kable(
      kable(umap_comp[[umap_name]], row.names = FALSE, format = "html") %>%
        kable_styling(bootstrap_options = c("striped", "condensed"),
                      full_width = FALSE, font_size = 12)
    )
    cat("\n\n")
  }

  # ===========================================================================
  # GLMM RESULTS — RAW FEATURES
  # ===========================================================================
  cat("## GLMM Results\n\n")
  cat("### Raw Features\n\n")

  lambda_index <- raw_glmm$optimal_lambda
  actual_lambda <- lambdas[lambda_index]

  cat("- **Optimal lambda:**", actual_lambda,
      paste0("(index ", lambda_index, " of ", length(lambdas), ")"), "\n")
  cat("- **CV RMSE:**", round(raw_glmm$rmse, 4), "\n\n")

  # ----- Coefficient table -----
  coef_df <- format_glmm_coefs(raw_glmm, "Raw")
  if (!is.null(coef_df)) {
    coef_df$Category <- classify_glmm_terms(coef_df$Term, predictor)

    nonzero <- coef_df %>% filter(Estimate != 0)
    n_zero <- nrow(coef_df) - nrow(nonzero)

    cat_order <- c("Intercept", "Focal predictor",
                   "Covariate main effect", "Interaction with focal predictor")
    nonzero$Category <- factor(nonzero$Category, levels = cat_order)
    nonzero <- nonzero %>% arrange(Category, desc(abs(Estimate)))

    focal_rows <- which(nonzero$Category %in%
                          c("Focal predictor", "Interaction with focal predictor"))

    cat("**Non-zero coefficients** (focal predictor terms highlighted):\n\n")
    print_kable(
      kable(nonzero, row.names = FALSE, format = "html") %>%
        kable_styling(bootstrap_options = c("striped", "condensed"),
                      full_width = FALSE, font_size = 12) %>%
        row_spec(focal_rows, bold = TRUE, background = "#FFF3CD")
    )
    cat("\n\n*", n_zero, "coefficients penalized to zero.*\n\n")
  }

  # ----- Predicted vs Observed (resubstitution) -----
  cat("#### Predicted vs. Observed\n\n")

  final_model_glmm <- raw_glmm$final_model
  glmm_preds <- get_glmm_fitted(final_model_glmm)
  glmm_actual <- y_var

  if (is.null(glmm_preds)) {
    cat("*Could not extract fitted values from glmmLasso model.*\n\n")
  } else {

    glmm_pred_df <- data.frame(Predicted = glmm_preds, Observed = glmm_actual)

    p_predobs_glmm <- ggplot(glmm_pred_df, aes(x = Observed, y = Predicted)) +
      geom_point(alpha = 0.3, size = 1) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
      geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 0.7) +
      labs(title = paste0("GLMM Predicted vs. Observed: ",
                          predictor, " -> ", params$outcome_label))
    print(p_predobs_glmm)
    save_plot(p_predobs_glmm, paste0(fig_prefix, "_glmm_pred_vs_obs"))
    cat("\n\n")

    # ----- Residual diagnostics -----
    cat("#### Residual Diagnostics\n\n")

    glmm_resids <- glmm_actual - glmm_preds
    resid_df <- data.frame(Fitted = glmm_preds, Residuals = glmm_resids)

    p_resid_fitted <- ggplot(resid_df, aes(x = Fitted, y = Residuals)) +
      geom_point(alpha = 0.3, size = 1) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
      geom_smooth(method = "loess", se = FALSE, color = "steelblue", linewidth = 0.7) +
      labs(title = paste0("GLMM Residuals vs. Fitted: ",
                          predictor, " -> ", params$outcome_label))
    print(p_resid_fitted)
    save_plot(p_resid_fitted, paste0(fig_prefix, "_glmm_resid_vs_fitted"))
    cat("\n\n")

    p_qq <- ggplot(resid_df, aes(sample = Residuals)) +
      stat_qq(alpha = 0.3, size = 1) +
      stat_qq_line(color = "red") +
      labs(title = paste0("GLMM Q-Q Plot: ",
                          predictor, " -> ", params$outcome_label),
           x = "Theoretical Quantiles", y = "Sample Quantiles")
    print(p_qq)
    save_plot(p_qq, paste0(fig_prefix, "_glmm_qqplot"))
    cat("\n\n")
  }

  # ----- PCA & UMAP GLMM results -----
  reordered <- reorder_by_variant(all_formulas, all_glmms)

  for (i in seq_along(reordered$results)) {
    variant <- get_variant_label(reordered$formulas[[i]])
    if (variant == "Raw") next

    glmm_i <- reordered$results[[i]]
    lambda_index_i <- glmm_i$optimal_lambda
    actual_lambda_i <- lambdas[lambda_index_i]

    cat("### ", variant, " Features\n\n", sep = "")
    cat("- **Optimal lambda:**", actual_lambda_i,
        paste0("(index ", lambda_index_i, " of ", length(lambdas), ")"), "\n")
    cat("- **CV RMSE:**", round(glmm_i$rmse, 4), "\n\n")

    coef_df_i <- format_glmm_coefs(glmm_i, variant)
    if (!is.null(coef_df_i)) {
      nonzero_i <- coef_df_i %>% filter(Estimate != 0)
      n_zero_i <- nrow(coef_df_i) - nrow(nonzero_i)
      cat("**Non-zero coefficients:**\n\n")
      print_kable(
        kable(nonzero_i, row.names = FALSE, format = "html") %>%
          kable_styling(bootstrap_options = c("striped", "condensed"),
                        full_width = FALSE, font_size = 12)
      )
      cat("\n\n*", n_zero_i, "coefficients penalized to zero.*\n\n")
    }
  }

  # ===========================================================================
  # RANDOM FOREST RESULTS (Raw only)
  # ===========================================================================
  cat("## Random Forest Results (Raw)\n\n")

  rf_model <- tryCatch({
    refit_ranger_with_importance(raw_forest, df_analysis, raw_formula)
  }, error = function(e) {
    cat("*Could not refit ranger model:", conditionMessage(e), "*\n\n")
    NULL
  })

  if (!is.null(rf_model)) {

    rec_pars <- raw_forest$recommended.pars
    cat("### Tuning Summary\n\n")
    cat("- **mtry:**", rec_pars$mtry, "\n")
    cat("- **min.node.size:**", rec_pars$min.node.size, "\n")
    cat("- **sample.fraction:**", round(rec_pars$sample.fraction, 3), "\n")
    cat("- **CV RMSE:**", round(raw_forest$rmse, 4), "\n\n")

    # ----- Variable importance (lollipop style) -----
    cat("### Variable Importance (Permutation)\n\n")

    imp <- importance(rf_model)
    imp_df <- data.frame(
      Variable = names(imp),
      Importance = imp,
      stringsAsFactors = FALSE
    ) %>%
      mutate(
        is_focal = Variable == predictor |
          grepl(paste0(":", predictor, "$|^", predictor, ":"), Variable),
        fill_group = ifelse(is_focal,
                            "Focal predictor (or interaction)", "Other")
      ) %>%
      arrange(desc(Importance))

    imp_top <- imp_df %>% slice_head(n = 25)
    imp_top$Variable <- ordered(imp_top$Variable, levels = rev(imp_top$Variable))

    p_imp <- ggplot(imp_top,
                    aes(y = Variable, x = Importance, color = fill_group)) +
      geom_segment(aes(x = 0, xend = Importance, y = Variable, yend = Variable),
                   colour = "grey50", linetype = 2) +
      geom_vline(xintercept = 0, colour = "grey50", linetype = 1) +
      geom_point(size = 2.5) +
      scale_color_manual(values = c(
        "Focal predictor (or interaction)" = "#E69F00",
        "Other" = "steelblue")) +
      labs(title = paste0("Variable Importance (Top 25): ",
                          predictor, " -> ", params$outcome_label),
           x = "Permutation Importance", color = NULL) +
      theme_bw() +
      theme(panel.grid.major.x = element_blank(),
            panel.grid.minor.x = element_blank(),
            axis.title.y = element_blank(),
            legend.position = "bottom")
    print(p_imp)
    save_plot(p_imp, paste0(fig_prefix, "_rf_importance"), height = 8)
    cat("\n\n")

    # Full importance table
    cat("**Full importance ranking:**\n\n")
    imp_table <- imp_df %>%
      select(Variable, Importance) %>%
      mutate(Importance = round(Importance, 6))
    focal_rows_imp <- which(imp_df$is_focal)
    print_kable(
      kable(imp_table, row.names = FALSE, format = "html") %>%
        kable_styling(bootstrap_options = c("striped", "condensed"),
                      full_width = FALSE, font_size = 11) %>%
        row_spec(focal_rows_imp, bold = TRUE, background = "#FFF3CD") %>%
        scroll_box(height = "400px")
    )
    cat("\n\n")

    # ----- Partial dependence plots -----
    cat("### Partial Dependence Plots\n\n")
    cat("PDP for the focal predictor and the top 5 most important ",
        "non-interaction covariates.\n\n", sep = "")

    pred_vars <- setdiff(all.vars(raw_formula), outcome_var)
    model_data <- df_analysis[, c(outcome_var, pred_vars)]

    # Focal predictor
    cat("#### Focal Predictor: ", predictor, "\n\n", sep = "")
    pdp_focal <- compute_pdp(rf_model, model_data, predictor)

    p_pdp_focal <- ggplot(pdp_focal, aes(x = value, y = yhat)) +
      geom_line(color = "#E69F00", linewidth = 1.2) +
      geom_point(color = "#E69F00", size = 2) +
      coord_cartesian(ylim = y_range) +
      labs(title = paste0("PDP: ", predictor, " -> ", params$outcome_label),
           x = predictor, y = paste0("Predicted ", out)) +
      theme(plot.title = element_text(face = "bold"))
    print(p_pdp_focal)
    save_plot(p_pdp_focal, paste0(fig_prefix, "_rf_pdp_", predictor))
    cat("\n\n")

    # Top 5 non-interaction covariates
    top5_vars <- imp_df %>%
      filter(Variable != predictor, !grepl(":", Variable)) %>%
      slice_head(n = 5) %>%
      pull(Variable)

    for (v in top5_vars) {
      pdp_v <- compute_pdp(rf_model, model_data, v)
      p_v <- ggplot(pdp_v, aes(x = value, y = yhat)) +
        geom_line(color = "steelblue", linewidth = 0.9) +
        geom_point(color = "steelblue", size = 1.5) +
        coord_cartesian(ylim = y_range) +
        labs(title = paste0("PDP: ", v, " (", predictor, " -> ",
                            params$outcome_label, ")"),
             x = v, y = paste0("Predicted ", out))
      print(p_v)
      save_plot(p_v, paste0(fig_prefix, "_rf_pdp_",
                            gsub("[^A-Za-z0-9_]", "_", v)))
      cat("\n\n")
    }

    # ----- Forest: Predicted vs Observed -----
    cat("### Predicted vs. Observed\n\n")

    rf_preds <- predict(rf_model, data = model_data)$predictions
    rf_actual <- model_data[[outcome_var]]

    rf_pred_df <- data.frame(Predicted = rf_preds, Observed = rf_actual)

    p_predobs_rf <- ggplot(rf_pred_df, aes(x = Observed, y = Predicted)) +
      geom_point(alpha = 0.3, size = 1) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
      geom_smooth(method = "lm", se = FALSE, color = "steelblue", linewidth = 0.7) +
      labs(title = paste0("Forest Predicted vs. Observed: ",
                          predictor, " -> ", params$outcome_label))
    print(p_predobs_rf)
    save_plot(p_predobs_rf, paste0(fig_prefix, "_rf_pred_vs_obs"))
    cat("\n\n")

  } # end if rf_model not null

  # ===========================================================================
  # MODEL COMPARISON
  # ===========================================================================
  cat("## Model Comparison\n\n")

  rmse_table <- tab_fits %>%
    mutate(
      feature_set = case_when(
        grepl("PCA", model)  ~ "PCA",
        grepl("UMAP", model) ~ "UMAP",
        TRUE ~ "Raw"),
      rmse = round(rmse, 4)) %>%
    select(method, feature_set, rmse) %>%
    tidyr::pivot_wider(names_from = feature_set, values_from = rmse) %>%
    arrange(match(method, c("glmm", "forest", "tree"))) %>%
    select(method, Raw, PCA, UMAP)

  cat("### RMSE Comparison (all methods and feature sets)\n\n")
  print_kable(
    kable(rmse_table, format = "html",
          col.names = c("Method", "Raw", "PCA", "UMAP")) %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                    full_width = FALSE) %>%
      column_spec(1, bold = TRUE)
  )
  cat("\n\n")

  cat("---\n\n")
}
```
